Synthesizing Quotient Lenses

### 2018-03-14 papers/QuotientLenses/main.tex
intro edits

In ~\cite{optician}, we developed a tool called {\em Optician} to synthesize the
class of {\em bijective lenses} that can be derived in the lens programming
language {\em Boomerang}\cite{boomerang}. Each lens $\ell : S \Leftrightarrow T$
that Optician can synthesized specifies a bijection from the language of a
source regular expression $S$ to a target language of a target regular
expression $T$: \begin{equation}\label{bijectivelenslaws} \ell.\get \;
(\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s) = s
\end{equation}
Optician makes programming bijective lenses faster and easier, and though
many useful lenses are not bijective in nature, they do have a bijective core
surrounded by structured transformations, hence Optician can be used as a
drop-in component for synthesizing more complex lenses.

This paper applies this idea to synthesize {\em quotient lenses}
~\cite{quotientlenses}. Quotient lenses are lenses in which the lens laws are
loosened so that they hold modulo an equivalence relation on the source and
target data respectively; in this paper we are concerned with {\em bijective
quotient lenses} which are lenses for which Equation \ref{bijectivelenslaws}
holds modulo equivalence relations $\sim_S$ and $\sim_T$ defined on the source
and target data respectively:
\begin{equation}\label{quotientlenslaws}
\ell.\get \; (\ell.\lput \; t) \sim_T t \text{, and } \ell.\lput \; (\ell.\get
\; s) \sim_S s
\end{equation}
Quotient lenses are useful in situations where a programmer wishes for the
transformation defined by a lens to have the same behavior on data that differ
only in inessential details. For instance, a programmer may wish to ``quotient
away'' the number of white space characters between data items, the
capitalization of various strings, the sequence of fields in a record, or the
order of items in a list.

Optician synthesizes bijective lenses from a pair $(S, T)$ of regular expression
specifying the source and target types and a set of example input-output pairs
that guide the synthesis algorithm. This presents a challenge for synthesizing
quotient lenses since a specification of the source and target formats for a
quotient lens needs to account for the equivalence relations defined on the
respective formats. Our solution is to introduce {\em Quotient Regular
Expressions} (QREs), which are regular expressions augmented with extra
syntax that enables programmers to simultaneously specify a regular expression
as well as an equivalence relation on the language of that regular expression.
Further, given a QRE, we can automatically infer a set of canonical
representatives the equivalence relation defined by the QRE as well as a
canonizing function to the canonical set of representatives.

@@

In summary, our main contributions are:
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs), 
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We introduce {\em QRE lenses}, which translate between data formats
  specified using QREs (Section~\ref{QRE-lenses}).
  \item We design a specific set of useful QRE lens combinators and prove
  that QRE lenses defined using these combinators can be put into a
  restricted normal form (Section~\ref{QRE-lenses}).  This is
  the main technical contribution of this paper.
  \item We leverage this result to reduce the problem of {\em synthesizing} QRE
  lenses to the problem of synthesizing bijective string lenses, which was
  previously studied in~\cite{optician} (Section~\ref{synth}).
  \item We extend Boomerang with QREs and QRE lens synthesis and demonstrate
  the utility and practicality of our approach by synthesizing QRE lenses
  between a variety of data formats drawn from the Optician benchmark suite
  (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} present related and future work.

>>>>

In recent work~\cite{optician}, we developed a tool called {\em Optician} to synthesize the
class of {\em bijective lenses} that can be derived in the lens programming
language {\em Boomerang}\cite{boomerang}. Each lens $\ell : S \Leftrightarrow T$
that Optician can synthesize specifies a bijection from the language of a
source regular expression $S$ to a target language of a target regular
expression $T$: \begin{equation}\label{bijectivelenslaws} \ell.\get \;
(\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s) = s
\end{equation}
Optician makes programming bijective lenses faster and easier, and though
many useful lenses are not bijective in nature, they do have a bijective core
surrounded by structured transformations. Hence, Optician can be used as a
component in a system that synthesizes more complex lenses.

This paper applies this idea to the problem of
synthesizing {\em quotient lenses}~\cite{quotientlenses}.
Quotient lenses are lenses in which the lens laws are
loosened so that they hold modulo an equivalence relation on the source and
target data respectively; in this paper we are concerned with {\em bijective
quotient lenses} which are lenses for which Equation \ref{bijectivelenslaws}
holds modulo equivalence relations $\sim_S$ and $\sim_T$ defined on the source
and target data respectively:
\begin{equation}\label{quotientlenslaws}
\ell.\get \; (\ell.\lput \; t) \sim_T t \text{, and } \ell.\lput \; (\ell.\get
\; s) \sim_S s
\end{equation}
Quotient lenses are useful in situations where a programmer wishes for the
transformation defined by a lens to have the same behavior on data that differ
only in inessential details. For instance, a programmer may wish to ``quotient
away'' the number of white space characters between data items, the
capitalization of various strings, the sequence of fields in a record, or the
order of items in a list.

Optician synthesizes bijective lenses from a pair $(S, T)$ of regular
expressions
specifying the source and target types and a set of example input-output pairs
that guide the synthesis algorithm. This presents a challenge for synthesizing
quotient lenses since a specification of the source and target formats for a
quotient lens needs to account for the equivalence relations defined on the
respective formats. Our solution is to introduce {\em Quotient Regular
Expressions} (QREs), which are regular expressions augmented with extra
syntax that enables programmers to simultaneously specify a regular expression
as well as an equivalence relation on the language of that regular expression.
Further, given a QRE, we can automatically infer a \emph{canonizer}---a
function that converts string in the language of regular expression
to a canonical form.

@@

In summary, our main contributions are:
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs), 
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We introduce {\em QRE lenses}, which translate between data formats
  specified using QREs (Section~\ref{QRE-lenses}).
  \item We design a specific set of useful QRE lens combinators and prove
  that QRE lenses defined using these combinators can be put into a
  restricted normal form (Section~\ref{QRE-lenses}).  This is
  the main technical contribution of this paper.
  \item We leverage this result to reduce the problem of {\em synthesizing} QRE
  lenses to the problem of synthesizing bijective string lenses, which was
  previously studied in past work~\cite{optician} (Section~\ref{synth}).
  \item We extend Boomerang with QREs and QRE lens synthesis and demonstrate
  the utility and practicality of our approach by synthesizing QRE lenses
  between a variety of data formats drawn from the Optician benchmark suite
  (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} present related and future work.
